Cryptobenthic crab assemblages are more distinct across a 90 m depth gradient than 2,500 km of shallow marine habitat in the Hawaiian Archipelago

### 2025-06-12 manuscript.Rmd
various revisions

---
title: 'A cryptobenthic reef assemblage is more distinct across a 90 m depth gradient than 2,500 km of shallow marine habitat in the Hawaiian Archipelago'
author:
  - Mykle L. Hoban:
      institute:
        - himb
      equal_contributor: yes
      correspondence: yes
      email: "mhoban@hawaii.edu"
  - Kaleonani K.C. Hurley:
      institute:
        - himb
      equal_contributor: yes
  - Kerry Reardon:
      institute: noaa
  - Derek J. Skillings:
      institute:
        - uncg
  - Molly A. Timmers:
      institute:
        - himb
        - natgeo
  - Robert J. Toonen:
      institute: himb
institute:
  - himb: Hawai&#x02BB;i Institute of Marine Biology, University of Hawai&#x02BB;i, Mānoa
  - noaa: National Oceanic and Atmospheric Administration
  - uncg: 'University of North Carolina, Greensboro'
  - natgeo: National Geographic Pristine Seas
bibliography: references/citations.json
csl: references/style.csl
output:
  word_document:
    pandoc_args:
      - "--filter=resources/pandoc-crossref"
      - "--citeproc"
      - "--toc-depth=2"
      - "--lua-filter=resources/scholarly-metadata.lua"
      - "--lua-filter=resources/author-info-blocks.lua"
      - "--lua-filter=resources/abstract-section.lua"
      - "--reference-doc=resources/ref.docx"
      - "--lua-filter=resources/multirefs.lua"
    keep_md: true
tblPrefix: Table
figPrefix:
  - "Fig."
  - "Figs."
chapters: True
chaptersDepth: 1
chapDelim: ""
lofTitle: |
  # Figure legends {-}
lofItemTitle: |
  Fig.
lofItemTemplate: |
  $$lofItemTitle$$ $$i$$$$listItemTitleDelim$$ $$t$$
header-includes:
  - \usepackage{bm}
  - \usepackage{upgreek}
multiref_no_duplicates: true
---
<!--word count:
Articles should ideally be no more than 11 typeset pages
The main text should be no more than 4,500 words (not including Abstract, Methods, References and figure legends)
The title should be no more than 20 words, should describe the main message of the article using a single scientifically accurate sentence, and should not contain puns or idioms
The abstract should be no more than 200 words
-->
### Keywords {-}
<div custom-style="mono">Mesophotic coral ecosystems, community structure, depth gradient, environmental drivers, Hawaii, Autonomous Reef Monitoring Structures</div>

@@

Investigations into the factors that shape coral reef communities have been historically limited in depth and taxonomic scope. Despite high biodiversity and the recognized importance of mesophotic habitats, and due to logistical challenges, studies have focused largely on conspicuous taxa such as fishes and corals and the shallowest ~30 m. Here, we examine the variability of crab assemblages from Autonomous Reef Monitoring Structures across Hawai&#x02BB;i and ~90 m depth. We examine the effects of environmental, ecological, and anthropogenic factors (latitude, longitude, depth, years, chlorophyll-A, temperature, slope, coral cover, potential larval immigration, human impacts) on shallow (8--17 m) crab assemblages and compare them with those sampled along a mesophotic depth gradient (12--90 m) on O&#x02BB;ahu. Shallow assemblages were significantly different between the inhabited Main and relatively pristine Northwestern Hawaiian Islands and associated with sea-surface temperature, chlorophyll-A, depth, slope, potential larval immigration, and human impacts. Despite this variability, differences along the depth gradient on O&#x02BB;ahu alone were greater than among the shallow assemblages across the entire archipelago. This finding suggests that 90 m of depth is a stronger determinant of brachyuran assemblage structure than the latitudinal, environmental, or anthropogenic gradients across the entire ~2,500 km span of the Hawaiian Islands.

@@

Coral reefs are among the most diverse habitats in the world, and they face increasing threats of degradation, including climate-driven bleaching and phase shifts to alternate ecological regimes [@grahamManagingResilienceReverse2013; @norstromAlternativeStatesCoral2009; @pandolfiGlobalTrajectoriesLongterm2003]. In the face of these threats, understanding drivers of metazoan community structure in coral reef ecosystems is critically important [@hughesRisingChallengeSustaining2010; @pageSeekingResistanceCoral2019], particularly as the search for criteria to prioritize climate refugia such as upwelling zones or mesophotic reefs intensifies. While the structure and composition of coral reef communities is influenced by a suite of environmental and anthropogenic factors, the nature and makeup of those factors remain poorly resolved [@hughesCommunityStructureDiversity1989; @smithEffectsTopBottom2010; @brandlCoralReefEcosystem2019]. One aspect of coral reef community structure thought to be important but remains under-investigated is depth. Most studies to date have focused on the upper ~30 m of the water column, yet coral reefs and associated communities are known to extend beyond 150 m [@pyleExploringDeepCoral1996; @pyleMesophoticCoralEcosystems2019], with photosymbiotic corals reported to 170 m [@rouzeSymbioticAssociationsDeepest2021]. As much as two thirds of coral reef ecosystems lie in this under-explored zone [@hindersteinThemeSectionMesophotic2010]. These habitats, known as mesophotic coral ecosystems (MCEs), are typically defined as those extending from ~30--150 m depth [@hindersteinThemeSectionMesophotic2010; @kahngCommunityEcologyMesophotic2010]. They have been hypothesized as possible refugia from climate change impacts on shallow reefs [@glynnCoralReefBleaching1996; @bongaertsAssessingDeepReef2010; @smithDepthRefugiumCatastrophic2014; @holsteinFertileFathomsDeep2015; @muirSpeciesIdentityDepth2017], although a growing number of studies have found evidence that mesophotic reefs are unique and uniquely threatened, challenging this notion [@diazLightTemperatureDrive2023; @diazMesophoticCoralBleaching2023; @hobanPlumbingDepthsEnvironmental2023; @hurleyAssessmentShallowMesophotic2016; @kaneHighLevelsMesophotic2014; @pinheiroDeepReefFishes2019; @rochaMesophoticCoralEcosystems2018; @montgomeryCommunitySimilaritySpecies2021].

The Hawaiian Archipelago, spanning ~2,500 km in the North Pacific Ocean, provides a natural laboratory to examine drivers of metazoan community structure. This chain of islands comprises two regions, the inhabited high islands of the main Hawaiian Islands (MHI) and the uninhabited low islands, atolls, and banks of the Northwestern Hawaiian Islands (NWHI) which reside within one of the world's largest marine protected areas---the Papahānaumokuākea Marine National Monument. The island chain is subject to considerable anthropogenic and physical gradients driving differences in temperature, coral cover, human population, and anthropogenic impact, among others, all of which influence coral reef communities [@selkoeMapHumanImpacts2009]. Nearshore drop-offs along the island chain also allow access to deeper reef habitat, enabling the examination of depth influences on community structure.

To date, the majority of studies that have examined community structure across the Hawaiian archipelago and/or along a depth gradient have focused on conspicuous taxa such as corals and fishes [e.g., @benfieldComparisonCoralReef2008; @caballero-aragonWaveExposureTemperature2023; @kaneTrophicDesignationLive2017; @langeWaveExposureShapes2021; @vanlierHabitatConnectivityComplexity2018]. One previous study [@hurleyAssessmentShallowMesophotic2016] found a significant depth gradient on the island of O&#x02BB;ahu for brachyuran crabs. However, the most frequently-targeted taxa make up a minority of coral reef biodiversity and thus may not be indicative of community patterns overall [@fautinOverviewMarineBiodiversity2010]. Furthermore, studies on these taxa have also been criticized for confounding multiple potentially covarying factors (such as habitat area and/or complexity) across the measured gradients [e.g., @bellwoodRegionalscaleAssemblyRules2001; @grahamEffectsMarineReserve2003; @connollyCommunityStructureCorals2005; @ewersConfoundingFactorsDetection2006; @fletcherRoleHabitatArea2007].

@@

Given the strong reported differences between deep and shallow reefs across taxa and the paucity of research on drivers of cryptic coral reef community variation, we sought to answer the following four research questions: (i) How do shallow reef brachyuran crab assemblages vary along the natural and anthropogenic gradients of the Hawaiian Archipelago? (ii) Do shallow reef brachyuran crab assemblages differ between the Main and Northwestern Hawaiian Islands? (iii) What environmental and anthropogenic factors drive crab assemblage structure in shallow coral reefs across the Hawaiian Archipelago? (iv) What are the relative contributions of depth (shallow vs. mesophotic) and island region to crab community variation across the Hawaiian Archipelago?

@@

everywhere <- cc$crabs_untransformed %>%
  psmelt() %>%
  filter(Abundance > 0) %>%
  group_by(OTU,species) %>%
  summarise(dp = n_distinct(depth_zone)) %>%
  filter(dp == ndepths) %>%
  arrange(species) %>%
  left_join(crab_taxonomy_validated,by=c("OTU" = "otu")) %>%
  mutate(display_species = as.character(case_when(
    as.logical(complex) ~ cf_species(display_species),
    .default = display_species
  ))) %>%
  pull(display_species)
```
Out of the `r all_summary$taxonomy_breakdown$families` families recovered, `r fam['shallow']` were exclusive to shallow ARMS recovered from `r dr(cc$crabs_shallow_untransformed)` m and `r fam['deep']` were exclusive to units $\geq$ 30 m (`r format_list(deep_fams)`; Supplementary [@tbl:tbl_supp_family_depth]). `r str_to_sentence(english(gen['shallow']))` genera were found exclusively shallow and `r gen['deep']` exclusively deep (`r format_list(deep_gens)`; Supplementary [@tbl:tbl_supp_genus_depth]). `r str_to_sentence(english(spp['shallow']))` distinct taxa (including those collapsed to genus or family) were exclusive to shallow ARMS units and `r spp['deep']` were found only deep (`r format_list(deep_spp)`; Supplementary [@tbl:tbl_supp_species_depth]). Out of the `r ntaxa(cc$crabs)` total recovered taxa, only `r length(everywhere)` were found across all `r ndepths` depth zones (`r format_list(everywhere)`). For a more in-depth taxonomic treatment of the assemblage composition of mesophotic vs. shallow ARMS units, see [@hurleyAssessmentShallowMesophotic2016].

@@

```{r fig-cap, fig.cap="CAP ordination of crab communities by island group and depth zone.", results="asis", fig.width=9, fig.height=6, dependson="cap-analysis"}
centroids <- pcap %>% group_by(sample_group) %>%
  summarise(LD1=mean(LD1),LD2=mean(LD2))
pcap <- pcap %>%
  mutate(
    oahu = factor(
      # if_else(island == "Oahu","oahu","noahu"),
      case_when(
        island == "Oahu" ~ "oahu",
        island == "French Frigate Shoals" ~ "ffs",
        .default = "noahu"
      ),
      levels = c("oahu","ffs","noahu")
    )
  )

@@

Two potential weaknesses of our dataset are the time periods over which ARMS units were deployed and the fact that mesophotic samples were obtained solely from the island of O&#x02BB;ahu. Our ARMS units were deployed across a broad range of years, but the majority were deployed during 2010--2012 or 2010--2013 (an overlap of two years) ([@tbl:tbl_arms]). A reduced version of the CAP ordination filtered to include only ARMS units deployed during these overlapping time periods retains the same (if not more pronounced) overall patterns as the full version (Supplementary [@fig:fig_supp_cap_reduced]), suggesting that the inclusion of prior or subsequent soaking periods did not adversely skew our analyses. Aditionally, soak time differed by one year between O&#x02BB;ahu mesophotic ARMS units and those deployed across the rest of the archipelago. This difference may potentially bias results, but since CAP ordinations show differences despite the extra year that shallow units spent in the water, we feel that the patterns we observed reflect a true separation. The limitation of our depth transect to O&#x02BB;ahu means that we cannot make broad-scale conclusions about biodiversity patterns on deep reefs across the Hawaiian Islands. However, it remains the case that community and alpha diversity differences among depth zones at that single island appear to overwhelm differences observed among shallow-water coral reef habitat across the archipelago and depth appears to be the strongest organizing factor in canonical space regardless of where or when the communities were sampled. Similar patterns were found at a local scale (tens of kilometers) on Hawai&#x02BB;i Island using environmental DNA (eDNA) [@hobanPlumbingDepthsEnvironmental2023]. In that study, communities were found to have more affinity within depth zones across sites than among depth zones within sites. Thus, despite the limitations in our dataset, the patterns we observed suggest that factors associated with the shallow-to-mesophotic depth gradient exert stronger ecological influence than those (including depth) that vary across the latitudinal, physical, and anthropogenic gradients throughout the Hawaiian Archipelago.

@@

A total of 89 ARMS units were deployed over the course of six years (2010--2016) in shallow (`r dr(cc$crabs_shallow_untransformed)` m depth) coral reefs in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:tbl_arms]). All units included in this study were deployed in forereef habitat and had a soaking period of three years (2010--2013 or 2013--2016). To explore the effects of depth, we also present data from an earlier study by Hurley et al. [-@hurleyAssessmentShallowMesophotic2016], in which 27 units were deployed off O&#x02BB;ahu from 2009--2012 in `r format_range(cc$crab_data %>% sample_tibble() %>% filter(island == "Oahu") %>% pull(depth))` m depth; nine shallow units were deployed at 12 m, and six each at depths of 30 m, 60 m, and 90 m ([@tbl:tbl_arms]). Shallow units were deployed according to NOAA-RAMP methodology, whereas units $\geq$ 30 m were deployed and recovered using technical mixed gas closed-circuit rebreather diving (see [@hurleyAssessmentShallowMesophotic2016]). Units from this study soaked for two years during 2009--2011 or 2010--2012. Except when discussing specific depth zones, we will refer to any units deployed at `r dr(cc$crabs_shallow_untransformed)` m as "shallow" and those deployed $\geq$ 30 m as "deep".

>>>>

---
title: 'Cryptobenthic crab assemblages are more distinct across a 90 m depth gradient than 2,500 km of shallow marine habitat in the Hawaiian Archipelago'
author:
  - Mykle L. Hoban:
      institute:
        - himb
      equal_contributor: yes
      correspondence: yes
      email: "mhoban@hawaii.edu"
  - Kaleonani K.C. Hurley:
      institute:
        - himb
      equal_contributor: yes
  - Kerry Reardon:
      institute: noaa
  - Derek J. Skillings:
      institute:
        - uncg
  - Molly A. Timmers:
      institute:
        - himb
        - natgeo
  - Robert J. Toonen:
      institute: himb
institute:
  - himb: Hawai&#x02BB;i Institute of Marine Biology, University of Hawai&#x02BB;i, Mānoa
  - noaa: National Oceanic and Atmospheric Administration
  - uncg: 'University of North Carolina, Greensboro'
  - natgeo: National Geographic Pristine Seas
bibliography: references/citations.json
csl: references/style.csl
output:
  word_document:
    pandoc_args:
      - "--filter=resources/pandoc-crossref"
      - "--citeproc"
      - "--toc-depth=2"
      - "--lua-filter=resources/scholarly-metadata.lua"
      - "--lua-filter=resources/author-info-blocks.lua"
      - "--lua-filter=resources/abstract-section.lua"
      - "--reference-doc=resources/ref.docx"
      - "--lua-filter=resources/multirefs.lua"
    keep_md: true
tblPrefix: Table
figPrefix:
  - "Fig."
  - "Figs."
chapters: True
chaptersDepth: 1
chapDelim: ""
lofTitle: |
  # Figure legends {-}
lofItemTitle: |
  Fig.
lofItemTemplate: |
  $$lofItemTitle$$ $$i$$$$listItemTitleDelim$$ $$t$$
header-includes:
  - \usepackage{bm}
  - \usepackage{upgreek}
multiref_no_duplicates: true
---
<!-- TODO: specify depth zones shallow, 30m, 60m, 90 m -->
<!--word count:
Articles should ideally be no more than 11 typeset pages
The main text should be no more than 4,500 words (not including Abstract, Methods, References and figure legends)
The title should be no more than 20 words, should describe the main message of the article using a single scientifically accurate sentence, and should not contain puns or idioms
The abstract should be no more than 200 words
-->
### Keywords {-}
<div custom-style="mono">Mesophotic coral ecosystems, community structure, depth gradient, environmental drivers, Hawaii, Autonomous Reef Monitoring Structures</div>

@@

Investigations into factors that shape coral reef communities have been historically limited in depth and taxonomic scope. Despite high biodiversity and the recognized importance of mesophotic habitats, and due to logistical challenges, studies have focused largely on conspicuous taxa such as fishes and corals and shallow habitat &lt;30 m. Here, we examined the variability of crab assemblages from Autonomous Reef Monitoring Structures deployed on shallow reefs across the Hawaiian Islands and a mesophotic depth gradient on O&#x02BB;ahu. We tested the effects of environmental, ecological, and anthropogenic factors<!-- (latitude, longitude, depth, years, chlorophyll-A, temperature, slope, coral cover, potential larval immigration, human impacts)--> on shallow (8--17 m) crab assemblages. We additionally compared shallow assemblages to those sampled along a depth gradient (12--90 m) on O&#x02BB;ahu. Shallow assemblages were significantly different between the inhabited Main and the relatively pristine uninhabited Northwestern Hawaiian Islands and were associated with sea-surface temperature, chlorophyll-A, depth, island slope, potential larval immigration, and human impacts. Despite this variability, differences along the depth gradient on O&#x02BB;ahu alone were greater than among the shallow assemblages across the entire archipelago. This finding suggests that 90 m of depth is a stronger determinant of brachyuran assemblage structure than the latitudinal, environmental, or anthropogenic gradients across the entire ~2,500 km span of the Hawaiian Islands.

@@

Coral reefs are among the most diverse habitats in the world, and they face increasing threats of degradation, including climate-driven bleaching and phase shifts to alternate ecological regimes [@grahamManagingResilienceReverse2013; @norstromAlternativeStatesCoral2009; @pandolfiGlobalTrajectoriesLongterm2003]. In the face of these threats, understanding drivers of metazoan community structure in coral reef ecosystems is critically important [@hughesRisingChallengeSustaining2010; @pageSeekingResistanceCoral2019], particularly as the search for criteria to prioritize climate refugia such as upwelling zones or mesophotic reefs intensifies. While the structure and composition of coral reef communities is influenced by a suite of environmental and anthropogenic factors, the nature and makeup of those factors remain poorly resolved [@hughesCommunityStructureDiversity1989; @smithEffectsTopBottom2010; @brandlCoralReefEcosystem2019]. One aspect of coral reef community structure thought to be important that remains under-investigated is depth (although interest in this area has grown considerably in recent years [@radiceRecentTrendsBiases2024]). Many studies to date have focused on the upper ~30 m of the water column, yet coral reefs and associated communities are known to extend beyond 150 m [@pyleExploringDeepCoral1996; @pyleMesophoticCoralEcosystems2019], with photosymbiotic corals reported to 170 m [@rouzeSymbioticAssociationsDeepest2021]. As much as two thirds of coral reef ecosystems lie in this under-explored zone [@hindersteinThemeSectionMesophotic2010]. These habitats, known as mesophotic coral ecosystems (MCEs), are typically defined as those extending from ~30--150 m depth [@hindersteinThemeSectionMesophotic2010; @kahngCommunityEcologyMesophotic2010]. They have been hypothesized as possible refugia from climate change impacts on shallow reefs [@glynnCoralReefBleaching1996; @bongaertsAssessingDeepReef2010; @smithDepthRefugiumCatastrophic2014; @holsteinFertileFathomsDeep2015; @muirSpeciesIdentityDepth2017], although a growing number of studies have found evidence that MCEs are unique and uniquely threatened, challenging this notion [@diazLightTemperatureDrive2023; @diazMesophoticCoralBleaching2023; @hobanPlumbingDepthsEnvironmental2023; @hurleyAssessmentShallowMesophotic2016; @kaneHighLevelsMesophotic2014; @pinheiroDeepReefFishes2019; @rochaMesophoticCoralEcosystems2018; @montgomeryCommunitySimilaritySpecies2021].

The Hawaiian Archipelago, spanning ~2,500 km in the North Pacific Ocean, provides a natural laboratory to examine drivers of metazoan community structure. This chain of islands comprises two regions, the inhabited high islands of the main Hawaiian Islands (MHI) and the uninhabited low islands, atolls, and banks of the Northwestern Hawaiian Islands (NWHI) which reside within one of the world's largest marine protected areas---the Papahānaumokuākea Marine National Monument. The island chain is subject to considerable anthropogenic and physical gradients driving differences in temperature, coral cover, human population, and anthropogenic impact, among others, all of which influence coral reefs and associated communities [@selkoeMapHumanImpacts2009]. Nearshore drop-offs along the island chain also allow access to deeper reef habitat, enabling the examination of depth influences on community structure.

To date, the majority of studies that have examined community structure across the Hawaiian archipelago and/or along a depth gradient have focused on conspicuous taxa such as corals and fishes [e.g., @benfieldComparisonCoralReef2008; @caballero-aragonWaveExposureTemperature2023; @kaneTrophicDesignationLive2017; @langeWaveExposureShapes2021; @vanlierHabitatConnectivityComplexity2018]. One previous study [@hurleyAssessmentShallowMesophotic2016] found a significant depth gradient on the island of O&#x02BB;ahu for brachyuran crabs. However, the most frequently-targeted taxa make up a minority of coral reef biodiversity and thus may not be indicative of community patterns overall [@fautinOverviewMarineBiodiversity2010]. Furthermore, some studies examining environmental effects on reef assemblages have been criticized for confounding multiple potentially covarying factors (such as habitat area and/or complexity) across the measured gradients [e.g., @bellwoodRegionalscaleAssemblyRules2001; @grahamEffectsMarineReserve2003; @connollyCommunityStructureCorals2005; @ewersConfoundingFactorsDetection2006; @fletcherRoleHabitatArea2007].

@@

Given the strong reported differences between deep and shallow reefs across taxa and the paucity of research on drivers of cryptic coral reef community variation, we sought to answer the following four research questions: (i) How do shallow reef brachyuran crab assemblages vary in response to the natural and anthropogenic gradients of the Hawaiian Archipelago? (ii) Do shallow reef brachyuran crab assemblages differ between the Main and Northwestern Hawaiian Islands? (iii) What environmental and anthropogenic factors drive crab assemblage structure in shallow coral reefs across the Hawaiian Archipelago? (iv) What are the relative contributions of depth (shallow vs. mesophotic) and island region to crab community variation across the Hawaiian Archipelago?

@@

dz <- cc$crab_data %>%
  distinct(depth_zone,depth) %>%
  mutate(depth_zone = fct_reorder(depth_zone,depth)) %>%
  pull(depth_zone) %>%
  levels()

everywhere <- cc$crabs_untransformed %>%
  psmelt() %>%
  filter(Abundance > 0) %>%
  group_by(OTU,species) %>%
  summarise(dp = n_distinct(depth_zone)) %>%
  filter(dp == ndepths) %>%
  arrange(species) %>%
  left_join(crab_taxonomy_validated,by=c("OTU" = "otu")) %>%
  mutate(display_species = as.character(case_when(
    as.logical(complex) ~ cf_species(display_species),
    .default = display_species
  ))) %>%
  pull(display_species)
```
Out of the `r all_summary$taxonomy_breakdown$families` families recovered, `r fam['shallow']` were exclusive to shallow ARMS recovered from `r dr(cc$crabs_shallow_untransformed)` m and `r fam['deep']` were exclusive to units $\geq$ 30 m (`r format_list(deep_fams)`; Supplementary [@tbl:tbl_supp_family_depth]). `r str_to_sentence(english(gen['shallow']))` genera were found exclusively shallow and `r gen['deep']` exclusively deep (`r format_list(deep_gens)`; Supplementary [@tbl:tbl_supp_genus_depth]). `r str_to_sentence(english(spp['shallow']))` distinct taxa (including those collapsed to genus or family) were exclusive to shallow ARMS units and `r spp['deep']` were found only deep (`r format_list(deep_spp)`; Supplementary [@tbl:tbl_supp_species_depth]). Out of the `r ntaxa(cc$crabs)` total recovered taxa, only `r length(everywhere)` were found across all `r ndepths` depth zones (`r format_list(dz)`; `r format_list(everywhere)`). For a more in-depth taxonomic treatment of the assemblage composition of mesophotic vs. shallow ARMS units, see [@hurleyAssessmentShallowMesophotic2016].

@@

```{r fig-cap, fig.cap="Canonical analysis of principal coordinates (CAP) ordination of crab assemblages by island group and depth zone.", results="asis", fig.width=9, fig.height=6, dependson="cap-analysis"}
centroids <- pcap %>% group_by(sample_group) %>%
  summarise(LD1=mean(LD1),LD2=mean(LD2))
pcap <- pcap %>%
  mutate(
    oahu = factor(
      # if_else(island == "Oahu","oahu","noahu"),
      case_when(
        island == "Oahu" ~ "oahu",
        island == "French Frigate Shoals" ~ "ffs",
        .default = "noahu"
      ),
      levels = c("oahu","ffs","noahu")
    )
  )

@@

Two potential weaknesses of our dataset are the time periods over which ARMS units were deployed and the fact that mesophotic samples were obtained solely from the island of O&#x02BB;ahu. Our ARMS units were deployed across a broad range of years, but the majority were deployed during 2010--2012 or 2010--2013 (an overlap of two years) ([@tbl:tbl_arms]). A reduced version of the CAP ordination filtered to include only ARMS units deployed during these overlapping time periods retains the same (if not more pronounced) overall patterns as the full version (Supplementary [@fig:fig_supp_cap_reduced]), suggesting that the inclusion of prior or subsequent soaking periods did not adversely skew our analyses. Additionally, soak time differed by one year between O&#x02BB;ahu mesophotic ARMS units and those deployed across the rest of the archipelago. This difference may potentially bias results, but since CAP ordinations show differences despite the extra year that shallow units spent in the water, we feel that the patterns we observed reflect a true separation. The limitation of our depth transect to O&#x02BB;ahu means that we cannot make broad-scale conclusions about biodiversity patterns on deep reefs across the Hawaiian Islands. However, it remains the case that community and alpha diversity differences among depth zones at that single island appear to overwhelm differences observed among shallow-water coral reef habitat across the archipelago and depth appears to be the strongest organizing factor in canonical space regardless of where or when the communities were sampled. Similar patterns were found at a local scale (tens of kilometers) on Hawai&#x02BB;i Island using environmental DNA (eDNA) [@hobanPlumbingDepthsEnvironmental2023]. In that study, communities were found to have more affinity within depth zones across sites than among depth zones within sites. Thus, despite the limitations in our dataset, the patterns we observed suggest that factors associated with the shallow-to-mesophotic depth gradient exert stronger ecological influence than those (including depth) that vary across the latitudinal, physical, and anthropogenic gradients throughout the Hawaiian Archipelago.

@@

A total of 89 ARMS units were deployed over the course of six years (2010--2016) in shallow (`r dr(cc$crabs_shallow_untransformed)` m depth) coral reefs in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:tbl_arms]). All shallow units included in this study were deployed in forereef habitat and had a soaking period of three years (2010--2013 or 2013--2016). To explore the effects of depth, we also present data from an earlier study by Hurley et al. [-@hurleyAssessmentShallowMesophotic2016], in which 27 units were deployed off O&#x02BB;ahu from 2009--2012 in `r format_range(cc$crab_data %>% sample_tibble() %>% filter(island == "Oahu") %>% pull(depth))` m depth; nine shallow units were deployed at 12 m, and six each at depths of 30 m, 60 m, and 90 m ([@tbl:tbl_arms]). Shallow units were deployed according to NOAA-RAMP methodology, whereas units $\geq$ 30 m were deployed and recovered using technical mixed gas closed-circuit rebreather diving (see [@hurleyAssessmentShallowMesophotic2016]). Units from this study soaked for two years during 2009--2011 or 2010--2012. Except when discussing specific depth zones, we will refer to any units deployed at `r dr(cc$crabs_shallow_untransformed)` m as "shallow" and those deployed $\geq$ 30 m as "deep".
